Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-01-22 ms.qmd
Fxing up Alex's comments

Climate change is expected to result in warmer but also more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses, high extinction rates are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates by buying time for genetic adaptation to take place [i.e., the 'plasticity first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity) or life-history stages.

@@

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @Pinsky2019-fn; @hoffmann2013upper]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022], it is unclear if the magnitude of physiological adjustment is sufficient, particularly in combination with potential changes in other factors, such as genetic and phenotypic variability. Low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both. Changes in physiological rate variance may therefore have important cascading effects on energy transfer and productivity for populations across these different ecosystems that are, as of yet, poorly measured and modelled.

>>>>

Climate change is expected to result in warmer but also more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will each contribute to a populations capacity to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses, high extinction rates are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates by buying time for genetic adaptation to take place [i.e., the 'plasticity first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity) or life-history stages.

@@

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to uncertainty over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper], whereas greater proximity to upper thermal limits in marine systems have questioned such conclusions [@Pinsky2019-fn]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022], it is unclear if the magnitude of physiological adjustment is sufficient, particularly in combination with potential changes in other factors, such as genetic and phenotypic variability. Low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both. Changes in physiological rate variance may therefore have important cascading effects on energy transfer and productivity for populations across these different ecosystems that are, as of yet, poorly measured and modelled.
